---
title: "Okra fruit growth: from raw observations to biological rates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Okra fruit growth: from raw observations to biological rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
library(nlrFlow)
```

## Why two okra datasets are distributed

`okra_growth_raw` preserves the 530 individual observations from the source script. `okra_growth_means` is a deterministic time-specific summary and is useful for curve illustration, but it should not silently replace the individual observations when uncertainty depends on within-time variation.

```{r}
raw <- nl_data("okra_growth_raw")
means <- nl_data("okra_growth_means")
nl_audit(raw, "fruit_length", "day_after_flowering")
nl_audit(means, "fruit_length", "day_after_flowering")
```

## Candidate growth equations

A first analysis compares several biologically plausible sigmoidal curves without treating non-nested equations as if they were nested models.

```{r}
fits <- nl_fit_many(
  means,
  response = "fruit_length",
  predictor = "day_after_flowering",
  models = c("logistic", "gompertz", "richards", "chapman_richards", "weibull_growth"),
  engine = "nls"
)
nl_compare(fits)
best <- nl_select(fits)
```

AIC/AICc/BIC and prediction metrics are evidence for model comparison, not proof that a curve is biologically correct. Use `nl_lrt()` only when mathematical nesting has been established explicitly.

## Diagnostics and parameter identifiability

```{r}
nl_diagnose(best)
nl_identify(best)
# Joint Beale confidence region for an nls-compatible fit:
# nl_confregion(best, n = 2000)
# Exact RMS curvature diagnostics are available through the optional IPEC backend.
# nl_curvature(best)
nl_influence(best)
nl_plot_diagnostics(best)
```

The identifiability step is particularly important for flexible four-parameter curves such as Richards. A visually close fit may still contain poorly separated parameters.

## Bootstrap uncertainty

```{r, eval=FALSE}
b <- nl_boot(best, R = 2000, type = "case", seed = 20260817)
nl_confint(best, method = "bootstrap", boot = b, bootstrap_interval = "bca")
nl_predict(best, interval = "confidence", nsim = 4000, seed = 20260817)
```

For inference from the raw individual observations, preserve their replication structure. A case bootstrap can resample observations, while a cluster bootstrap should be preferred when a real experimental-unit identifier exists.

## Absolute and relative fruit-growth rates

```{r}
der <- nl_derive(best)
der$summary
head(der$curve)
```

`AGR` approximates the first derivative of the fitted response. `RGR` divides that rate by the fitted response. The package also reports the inflection location, maximum absolute growth rate, its timing, the midpoint time and the area under the curve.

```{r, eval=FALSE}
der_b <- nl_derive(best, boot = b, level = 0.95)
nl_table(der_b, type = "derived", render = "gt")
```

## Resistant and quantile sensitivity analyses

The raw dataset contains substantial within-time dispersion. Resistant and nonlinear quantile fits can be useful sensitivity analyses when a mean Gaussian curve is not an adequate summary.

```{r, eval=FALSE}
f_rob <- nl_robust(
  fruit_length ~ Asym * exp(-exp(-k * (day_after_flowering - xmid))),
  raw,
  start = list(Asym = 18, k = 0.4, xmid = 4)
)
f_q50 <- nl_quantile(
  fruit_length ~ Asym * exp(-exp(-k * (day_after_flowering - xmid))),
  raw,
  start = list(Asym = 18, k = 0.4, xmid = 4), tau = 0.50
)
f_q90 <- nl_quantile(
  fruit_length ~ Asym * exp(-exp(-k * (day_after_flowering - xmid))),
  raw,
  start = list(Asym = 20, k = 0.4, xmid = 4), tau = 0.90
)
```

## Publication output

```{r, eval=FALSE}
p <- nl_plot(best, interval = "confidence", raw_data = raw)
nl_save_plot(p, "okra_growth_curve.tiff", width = 180, height = 120,
             units = "mm", dpi = 600)
nl_table(best, type = "parameters", render = "gt")
```

The raw observations should normally remain visible in the final scientific figure, even when the curve was initially selected with time-specific means for a didactic comparison.

# Bayesian alternatives can be diagnosed and compared after fitting:
# nl_bayes_diagnose(okra_bayes)
# nl_bayes_compare(list(Gompertz=okra_gomp_bayes, Richards=okra_rich_bayes))
