---
title: "Automatic Differentiation, RTMB, and Fast Bayesian Approximation"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Automatic Differentiation, RTMB, and Fast Bayesian Approximation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Scientific role

`nlrFlow` uses RTMB as an optional low-level engine when a nonlinear likelihood, latent-variable structure, or random-effect model does not fit a higher-level backend. RTMB keeps the statistical model written in R while recording derivatives by automatic differentiation. BayesRTMB can then route a compatible model to MAP/Laplace, NUTS, or ADVI.

The purpose is not to replace `nls()` for simple curves. Automatic differentiation becomes valuable when the likelihood itself is complex, when many latent effects are integrated, or when gradients from finite differences become unstable.

```{r rtmb-okra, eval=FALSE}
d <- subset(nl_data("agronomy_growth"), cultivar == "C1")
obj <- function(p) {
  mu <- p$A/(1 + exp(-p$k*(d$day-p$tm)))
  -sum(dnorm(d$biomass_Mg_ha, mu, exp(p$ls), log=TRUE))
}
fit_rtmb <- nl_rtmb(obj, list(A=18,k=.08,tm=50,ls=0))
der <- nl_ad(obj, list(A=18,k=.08,tm=50,ls=0))
```

The gradient returned by `nl_ad()` is generated by RTMB. Its optional Hessian is a local curvature calculation based on the AD gradient and is therefore labeled accordingly rather than presented as symbolic differentiation.

# BayesRTMB routing

```{r bayesrtmb, eval=FALSE}
# model <- BayesRTMB::rtmb_model(code, data = dat)
fit_map <- nl_bayes_rtmb(model, "optimize", laplace=TRUE)
fit_vb  <- nl_bayes_approx(model, "advi")
fit_lap <- nl_bayes_approx(model, "laplace")
```

ADVI is a computational approximation and can underestimate uncertainty. It is most useful as a screening or initialization route before full posterior simulation when the scientific decision depends on tail probabilities or interval width.

## Method references

RTMB builds on automatic differentiation and Laplace approximation [@Kristensen2016TMB]. Current RTMB/BayesRTMB software metadata are recorded in the package reference ledger.
