---
title: "Knowledge-Guided Symbolic Regression and Equation Discovery"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Knowledge-Guided Symbolic Regression and Equation Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Symbolic regression is a hypothesis generator

`nl_symbolic()` optionally calls PySR, which uses SymbolicRegression.jl. The returned equation set is deliberately not converted automatically into a biological conclusion. Complexity control and restricted operators reduce the search space, but scientific constraints remain the user's responsibility.

```{r pysr, eval=FALSE}
o <- nl_data("okra_growth_means")
sr_o <- nl_symbolic(o,"fruit_length","day_after_flowering",niterations=200,binary_operators=c("+","-","*","/"),unary_operators=c("exp"))

f <- nl_data("soil_fertility_p")
sr_f <- nl_symbolic(f,"grain_yield_Mg_ha","P2O5_kg_ha",niterations=200,binary_operators=c("+","-","*","/"),unary_operators=c("exp","log"))

p <- nl_data("plant_physiology_light")
sr_p <- nl_symbolic(p,"A_umol_CO2_m2_s","PAR_umol_m2_s",niterations=200,binary_operators=c("+","-","*","/"),unary_operators=c("sqrt"))
```

The recommended next step is to translate scientifically defensible expressions into interpretable parameterizations. `nl_discover()` then evaluates established mechanistic equations and new candidate formulas under the same inference and validation workflow.

```{r discover, eval=FALSE}
do <- nl_discover(o,"fruit_length","day_after_flowering",c("gompertz","logistic","richards"),constraints=list(positive=TRUE,monotonic="increasing"),engine="nls")
ff <- subset(nl_data("soil_fertility_p"),soil_class=="Loamy")
df <- nl_discover(ff,"grain_yield_Mg_ha","P2O5_kg_ha",c("mitscherlich","michaelis_menten","linear_plateau"),constraints=list(positive=TRUE),engine="nls")
ss <- subset(nl_data("soil_p_sorption"),soil_class=="Clayey")
ds <- nl_discover(ss,"sorbed_P_mg_kg","solution_P_mg_L",c("langmuir","freundlich"),constraints=list(positive=TRUE,monotonic="increasing"),engine="nls")
```

# Scientific validation before ranking

```{r validate, eval=FALSE}
vo <- nl_validate_candidate(do$fits$gompertz,"day_after_flowering",list(positive=TRUE,monotonic="increasing",response_range=c(0,30)))
vf <- nl_validate_candidate(df$fits$mitscherlich,"P2O5_kg_ha",list(positive=TRUE,monotonic="increasing"))
vs <- nl_validate_candidate(ds$fits$langmuir,"solution_P_mg_L",list(positive=TRUE,monotonic="increasing"))
```

A candidate can therefore fit the observed data well yet fail because it becomes negative, changes direction outside the observed support, violates plausible bounds, or is predictively unstable under cross-validation.

## Method references

PySR is described by @Cranmer2023. Knowledge-guided symbolic regression integrated with model-based experimental design is exemplified by @Rogers2024.
