---
title: "Universal Differential Equations for Mechanistic Agronomy"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Universal Differential Equations for Mechanistic Agronomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Why a UDE?

A Universal Differential Equation (UDE) preserves a known scientific rate law and adds a learnable term only where the mechanism is incomplete. This is the preferred Scientific Machine Learning option in `nlrFlow` whenever credible agronomic knowledge exists. The design follows the UDE principle of combining scientific models with universal approximators [@Rackauckas2020UDE; @Philipps2025UDEReview].

\[
\frac{d\mathbf y}{dt}=f_{known}(\mathbf y,\mathbf x,\theta)+f_{NN}(\mathbf y,\mathbf x,t;\phi).
\]

# 1. Crop growth: radiation-use-efficiency mechanism plus learned correction

The built-in `crop_growth_rue` template uses a known radiation/temperature/water response and lets the neural component capture remaining dynamics. The teaching data are synthetic and are designed for workflow validation, not for estimating universal agronomic constants.

```{r crop-ude, eval=FALSE}
crop <- nl_data("sciml_crop_growth")
p1 <- subset(crop, plot_id == unique(plot_id)[1])

ude_crop <- nl_ude(
  p1, time="day", states="biomass_g_m2",
  covariates=c("temperature_C","PAR_MJ_m2_d","soil_water_rel","nitrogen_rel"),
  template="crop_growth_rue",
  known_parameters=c(RUE=2.8, Topt=26, Twidth=10, respiration=.01),
  hidden=c(12,12), seed=20260817, maxiters=1500
)

nl_plot(ude_crop)
nl_table(ude_crop)
missing_crop <- nl_missing_physics(ude_crop, threshold=.25)
nl_plot(missing_crop)
nl_table(missing_crop)
```

The decomposition is the key output. A large neural correction is not automatically a success: it may indicate missing biology, a wrong known model, measurement error, or inadequate covariates.

# 2. Fruit growth: known saturation process plus environmental correction

```{r fruit-ude, eval=FALSE}
fruit <- nl_data("sciml_fruit_growth")
f1 <- subset(fruit, fruit_id == unique(fruit_id)[1])
ude_fruit <- nl_ude(
  f1, "day_after_set", "fruit_mass_g",
  c("temperature_C","soil_water_rel"),
  template="fruit_growth", known_parameters=c(r=.20,K=200),
  seed=20260818, maxiters=1200
)
nl_missing_physics(ude_fruit, threshold=.20)
nl_plot(ude_fruit)
```

# 3. Nitrogen uptake and fertility

```{r fertility-ude, eval=FALSE}
fert <- nl_data("sciml_fertility_dynamic")
f1 <- subset(fert, plot_id == unique(plot_id)[1])
ude_n <- nl_ude(
  f1, "day", c("plant_N_g_m2","soil_mineral_N_mg_kg"),
  c("temperature_C","soil_water_rel"),
  template="nitrogen_uptake", known_parameters=c(vmax=.08,Km=20),
  seed=20260819, maxiters=1600
)
nl_missing_physics(ude_n, threshold=.30)
nl_table(ude_n)
```

Here the scientific question is not simply whether the hybrid model predicts better. The relevant question is whether the learned correction is reproducible and structured enough to motivate a new nutrient-uptake hypothesis.

# Practical reporting

A publication-ready UDE analysis should report the known equations and parameters, neural architecture, solver, sensitivity method, optimizer sequence, seeds, training and validation split, learned-correction magnitude, extrapolation assessment, and an ablation comparison: mechanistic only versus UDE versus fully neural model.

# When not to use a UDE

Avoid a UDE if there is no credible known component, if the data are too sparse to identify both known and learned terms, or if the neural correction overwhelms the mechanistic component across the whole observed domain. In the last situation, reconsider the mechanistic model before interpreting the correction biologically.
