---
title: "Physics-Informed and Neural Dynamics in Plant Physiology"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Physics-Informed and Neural Dynamics in Plant Physiology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Why dynamic physiology?

Plant physiological observations frequently couple rapidly changing radiation, vapor-pressure deficit, temperature, stomatal conductance, and assimilation. A static nonlinear light-response curve remains preferable when it answers the question. Neural differential models become useful when the scientific target is the *dynamic adjustment rate* and environmental forcing changes through time.

Recent agricultural work has begun to use physics-informed models for plant-growth dynamics, including plant-height prediction [@Shao2026PlantHeight]. This vignette illustrates how `nlrFlow` keeps a conventional scientific comparator in the analysis rather than assuming the neural model is superior.

# Multistate Neural ODE

```{r node-phys, eval=FALSE}
phys <- nl_data("sciml_physiology_dynamic")
p1 <- subset(phys,plant_id==unique(plant_id)[1])
node <- nl_neural_ode(
  p1,"hour",c("A_umol_CO2_m2_s","gs_mol_H2O_m2_s"),
  c("PAR_umol_m2_s","VPD_kPa","leaf_temperature_C","soil_water_rel"),
  hidden=c(24,24),seed=20260817,maxiters=1400
)
nl_plot(node)
nl_sciml_diagnose(node)
```

# UDE when physiology is partially known

A preferred research formulation would keep a known light-response or gas-exchange component and let the neural term represent lagged stress effects.

```{r ude-phys, eval=FALSE}
known_rhs <- c(A_umol_CO2_m2_s="0.002*PAR_umol_m2_s - 0.05*A_umol_CO2_m2_s",
               gs_mol_H2O_m2_s="0.00015*PAR_umol_m2_s - 0.08*VPD_kPa*gs_mol_H2O_m2_s")
ude <- nl_ude(
  p1,"hour",c("A_umol_CO2_m2_s","gs_mol_H2O_m2_s"),
  c("PAR_umol_m2_s","VPD_kPa","leaf_temperature_C","soil_water_rel"),
  template="generic",known_rhs=known_rhs,seed=20260818,maxiters=1500
)
nl_missing_physics(ude,threshold=.20)
nl_plot(ude)
```

The equations in this teaching example are illustrative and are not presented as a validated physiological mechanism. In a research analysis, the known component must come from an appropriate physiological formulation and use consistent units.

# Simple PINN educational comparator

```{r pinn-phys, eval=FALSE}
# A low-dimensional constrained process before attempting a complex physiological PDE.
nl_pinn(p1,"logistic_growth","hour","A_umol_CO2_m2_s",parameters=c(r=.2,K=25),seed=20260819)
nl_pinn(p1,"first_order_nutrient","hour","gs_mol_H2O_m2_s",parameters=c(k=.1,K=.8),seed=20260820)
nl_pinn(p1,"logistic_growth","hour","gs_mol_H2O_m2_s",parameters=c(r=.15,K=.8),seed=20260821)
```

# Practical figures and tables

For physiological interpretation, publication figures should show observed trajectories, predicted trajectories, environmental forcing, and the learned correction. Tables should summarize dynamic prediction error and neural contribution rather than network weights. A neural correction peaking during high VPD, for example, is a hypothesis-generating result and requires independent validation before causal interpretation.
