---
title: "Missing-Physics Discovery in Soil Fertility and Soil Processes"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Missing-Physics Discovery in Soil Fertility and Soil Processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# From residual pattern to a process hypothesis

`nl_missing_physics()` is deliberately more conservative than naming the neural term a new mechanism. It decomposes an executed UDE into known, neural, and total derivative contributions and flags intervals where the neural fraction is substantial. The output is a *diagnostic for missing process representation*.

# 1. Nitrogen dynamics

```{r n, eval=FALSE}
fert <- nl_data("sciml_fertility_dynamic")
f1 <- subset(fert,plot_id==unique(plot_id)[1])
ude_n <- nl_ude(
  f1,"day",c("plant_N_g_m2","soil_mineral_N_mg_kg"),
  c("temperature_C","soil_water_rel"),template="nitrogen_uptake",
  known_parameters=c(vmax=.08,Km=20),seed=20260817,maxiters=1600
)
miss_n <- nl_missing_physics(ude_n,state="plant_N_g_m2",threshold=.25)
nl_plot(miss_n)
nl_table(miss_n)
```

Interpret high neural contribution in relation to temperature, water status, growth stage, and measurement support. It may represent omitted mineralization, root activity, immobilization, or simply misspecified known dynamics.

# 2. Crop biomass and nutrient limitation

```{r crop, eval=FALSE}
crop <- nl_data("sciml_crop_growth")
c1 <- subset(crop,plot_id==unique(plot_id)[1])
ude_c <- nl_ude(c1,"day","biomass_g_m2",
  c("temperature_C","PAR_MJ_m2_d","soil_water_rel","nitrogen_rel"),
  template="crop_growth_rue",known_parameters=c(RUE=2.8,Topt=26,Twidth=10,respiration=.01),seed=20260818)
miss_c <- nl_missing_physics(ude_c,state="biomass_g_m2",threshold=.20)
nl_plot(miss_c)
```

# 3. Soil organic carbon template

```{r soc, eval=FALSE}
# The soil_carbon template is available for a user-supplied longitudinal SOC dataset.
# Here dry_run documents the complete backend specification without pretending that
# the teaching package contains a validated long-term SOC experiment.
soc_demo <- data.frame(day=c(0,30,60,90),SOC=c(20,19.6,19.3,19.1),temperature_C=c(20,22,24,22),soil_water_rel=c(.65,.60,.55,.62))
ude_soc <- nl_ude(soc_demo,"day","SOC",c("temperature_C","soil_water_rel"),template="soil_carbon",known_parameters=c(k=.002,Q10=2),dry_run=TRUE)
# A decomposition becomes available after executing the SciML backend.
```

Recent work has explored UDEs for soil organic carbon, but the cited current example is a preprint and should be treated as emerging evidence rather than a settled agronomic standard [@Gress2025SOCUDE].

# Practical decision rule

A neural correction is scientifically interesting when it is: reproducible across seeds, concentrated in interpretable environmental or phenological regions, retained under validation, smaller than the total dynamics for much of the domain, and capable of being approximated by a parsimonious symbolic expression. If not, it is primarily evidence of model inadequacy, not evidence for a new biological law.
