---
title: "Getting Started with nlrFlow: A Practical Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with nlrFlow: A Practical Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
set.seed(20260817)
```

## Overview

`nlrFlow` provides an integrated workflow for nonlinear regression in agronomy and the natural sciences. This tutorial walks through a complete analysis: from data exploration to model fitting, diagnostics, comparison, and publication-ready outputs.

## 1. Load the package and explore data

```{r load-data}
library(nlrFlow)

# Okra fruit growth data
okra <- nl_data("okra_growth_means")
head(okra)
```

## 2. Audit the data

Before fitting, audit the structure:

```{r audit}
nl_audit(okra, "fruit_length", "day_after_flowering")
```

## 3. Generate starting values

`nl_start()` creates data-informed starting values for registered models:

```{r start-values}
nl_start(okra, "fruit_length", "day_after_flowering", "gompertz")
nl_start(okra, "fruit_length", "day_after_flowering", "logistic")
nl_start(okra, "fruit_length", "day_after_flowering", "richards")
```

## 4. Fit models

Fit the Gompertz model using Levenberg-Marquardt:

```{r fit-gompertz}
f_gomp <- nl_fit(
  data = okra,
  model = "gompertz",
  response = "fruit_length",
  predictor = "day_after_flowering",
  engine = "nlsLM"
)
summary(f_gomp)
```

Fit multiple candidate models at once:

```{r fit-many}
fits <- nl_fit_many(
  okra, "fruit_length", "day_after_flowering",
  c("logistic", "gompertz", "richards"),
  engine = "nlsLM"
)
```

## 5. Compare models

```{r compare}
nl_compare(fits)
```

## 6. Diagnostics

```{r diagnose}
nl_diagnose(f_gomp)
```

## 7. Predictions and confidence intervals

```{r predict}
pred <- nl_predict(f_gomp, interval = "confidence")
head(pred)
```

## 8. Derived quantities

Compute absolute growth rate, relative growth rate, inflection point, and area under the curve:

```{r derive}
d <- nl_derive(f_gomp)
d$summary
```

## 9. Plot

```{r plot, fig.cap="Gompertz fit to okra fruit growth"}
nl_plot(f_gomp)
```

## 10. Export a report

```{r report, eval=FALSE}
nl_report(f_gomp, file = "nlrFlow_report.md")
```

## Summary

This tutorial demonstrated the core `nlrFlow` workflow:

1. **Data audit** with `nl_audit()`
2. **Starting values** with `nl_start()`
3. **Model fitting** with `nl_fit()` and `nl_fit_many()`
4. **Model comparison** with `nl_compare()`
5. **Diagnostics** with `nl_diagnose()`
6. **Predictions** with `nl_predict()`
7. **Derived quantities** with `nl_derive()`
8. **Visualization** with `nl_plot()`
9. **Reporting** with `nl_report()`

For more details, see the other vignettes and the package documentation.

## How to cite

```{r citation, eval=FALSE}
citation("nlrFlow")
packageVersion("nlrFlow")
R.version.string
```
