---
title: "General agronomy: cultivar growth, qualitative factors, and hierarchical curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General agronomy: cultivar growth, qualitative factors, and hierarchical curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
library(nlrFlow)
```

## Experimental structure

The frozen teaching dataset represents biomass trajectories for three cultivars in blocks. It is simulated and exists to show how agronomic design information changes a nonlinear analysis.

```{r}
a <- nl_data("agronomy_growth")
a$unit_id <- interaction(a$cultivar, a$block, drop = TRUE)
nl_audit(a, "biomass_Mg_ha", "day", c("cultivar", "block"))
```

## Step 1: understand the curve family within one cultivar

```{r}
one <- subset(a, cultivar == "C1")
candidates <- nl_fit_many(
  one, "biomass_Mg_ha", "day",
  c("logistic", "gompertz", "richards", "weibull_growth", "von_bertalanffy"),
  engine = "nls"
)
nl_compare(candidates)
f1 <- nl_select(candidates)
nl_diagnose(f1)
nl_derive(f1)
```

This model-selection step is exploratory. Once a useful functional form has been chosen, cultivar effects should be tested in a joint model rather than by comparing significance labels from three independent fits.

## Step 2: qualitative cultivar effects on nonlinear parameters

```{r, eval=FALSE}
g <- nl_fit(
  biomass_Mg_ha ~ Asym/(1 + exp(-k * (day - xmid))),
  data = a,
  start = c(18, 2, 1, 0.09, -0.005, 0.01, 52),
  engine = "gnls",
  params = list(Asym ~ cultivar, k ~ cultivar, xmid ~ 1)
)
summary(g)
```

This specification asks whether cultivars differ in asymptotic biomass and growth-rate parameter while sharing the midpoint. Alternative nested parameter structures can be built deliberately and compared only when their nesting is genuine.

## Step 3: repeated observations and random effects

```{r, eval=FALSE}
m <- nl_mixed(
  biomass_Mg_ha ~ Asym/(1 + exp(-k * (day - xmid))),
  data = a,
  fixed = list(Asym ~ cultivar, k ~ cultivar, xmid ~ 1),
  random = Asym ~ 1 | unit_id,
  start = c(18, 2, 1, 0.09, -0.005, 0.01, 52),
  method = "ML"
)
summary(m)
```

The random effect represents heterogeneity among experimental units, not an arbitrary device to increase model complexity. More complex random structures should be justified by the design and by identifiability.

## Step 4: heteroscedastic or serially correlated trajectories

```{r, eval=FALSE}
m_varcor <- nl_mixed(
  biomass_Mg_ha ~ Asym/(1 + exp(-k * (day - xmid))),
  data = a,
  fixed = list(Asym ~ cultivar, k ~ cultivar, xmid ~ 1),
  random = Asym ~ 1 | unit_id,
  start = c(18, 2, 1, 0.09, -0.005, 0.01, 52),
  weights = nlme::varPower(form = ~ fitted(.)),
  correlation = nlme::corCAR1(form = ~ day | unit_id),
  method = "ML"
)
```

## Step 5: Bayesian multilevel extension

```{r, eval=FALSE}
bf <- brms::bf(
  biomass_Mg_ha ~ Asym/(1 + exp(-k * (day - xmid))),
  Asym ~ 1 + cultivar + (1 | unit_id),
  k ~ 1 + cultivar,
  xmid ~ 1,
  nl = TRUE
)
pri <- c(
  brms::prior(normal(18, 5), nlpar = "Asym"),
  brms::prior(lognormal(log(0.08), 0.5), nlpar = "k", lb = 0),
  brms::prior(normal(50, 15), nlpar = "xmid")
)
bm <- nl_bayes(bf, a, prior = pri, family = gaussian(), chains = 4, cores = 4)
nl_ppcheck(bm, type = "dens_overlay")
```

The Bayesian specification makes the parameter model and prior information explicit. Posterior probabilities and derived quantities can then be interpreted directly instead of reducing all cultivar comparisons to p-values.

## Publication workflow

```{r, eval=FALSE}
nl_table(f1, "parameters", "gt")
p <- nl_plot(f1, interval = "confidence")
nl_save_plot(p, "cultivar_growth.svg", width = 180, height = 120, units = "mm")
```

```{r eval=FALSE}
nl_bayes_diagnose(bm)
# With two scientifically plausible Bayesian models:
# nl_bayes_compare(list(Model1=bm, Model2=bm_alt), method="stacking")
```
