---
title: "Plant physiology: photosynthetic light response, water regimes, and nonlinear hierarchy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plant physiology: photosynthetic light response, water regimes, and nonlinear hierarchy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
library(nlrFlow)
```

## Light-response data

```{r}
p <- nl_data("plant_physiology_light")
nl_audit(p, "A_umol_CO2_m2_s", "PAR_umol_m2_s", c("water_regime", "plant_id"))
ww <- subset(p, water_regime == "WellWatered")
```

## Rectangular versus non-rectangular hyperbola

```{r}
f_rect <- nl_fit(
  data = ww, model = "rectangular_hyperbola",
  response = "A_umol_CO2_m2_s", predictor = "PAR_umol_m2_s", engine = "nlsLM",
  lower = c(Amax = 0, alpha = 1e-6, Rd = 0)
)
f_nonrect <- nl_fit(
  data = ww, model = "nonrectangular_hyperbola",
  response = "A_umol_CO2_m2_s", predictor = "PAR_umol_m2_s", engine = "nlsLM",
  lower = c(Amax = 0, alpha = 1e-6, theta = 0.05, Rd = 0),
  upper = c(Amax = Inf, alpha = Inf, theta = 0.999, Rd = Inf)
)
nl_compare(list(Rectangular = f_rect, Nonrectangular = f_nonrect))
nl_diagnose(f_nonrect)
nl_identify(f_nonrect)
# nl_confregion(f_nonrect, n = 2000)
# nl_curvature(f_nonrect)
```

The curvature parameter of the non-rectangular model can be weakly identified if the PAR range does not cover both low-light and saturating regions. Parameter constraints should reflect physiology, not simply optimizer convenience.

## Water regime as a qualitative factor

```{r, eval=FALSE}
g <- nl_fit(
  A_umol_CO2_m2_s ~ Amax*alpha*PAR_umol_m2_s/(Amax + alpha*PAR_umol_m2_s) - Rd,
  data = p,
  start = c(28, -5, -12, 0.07, -0.01, -0.02, 1.3),
  engine = "gnls",
  params = list(Amax ~ water_regime, alpha ~ water_regime, Rd ~ 1),
  weights = nlme::varPower(form = ~ pmax(abs(fitted(.)), 0.1))
)
summary(g)
```

## Plant-level nonlinear mixed model

```{r, eval=FALSE}
m <- nl_mixed(
  A_umol_CO2_m2_s ~ Amax*alpha*PAR_umol_m2_s/(Amax + alpha*PAR_umol_m2_s) - Rd,
  data = p,
  fixed = list(Amax ~ water_regime, alpha ~ water_regime, Rd ~ 1),
  random = Amax ~ 1 | plant_id,
  start = c(28, -5, -12, 0.07, -0.01, -0.02, 1.3),
  method = "ML"
)
```

This model preserves repeated measurements within plant. Treating every gas-exchange measurement as an independent replicate would overstate the effective sample size.

## Bayesian Student-t model for heavy-tailed measurement error

```{r, eval=FALSE}
bf <- brms::bf(
  A_umol_CO2_m2_s ~ Amax*alpha*PAR_umol_m2_s/(Amax + alpha*PAR_umol_m2_s) - Rd,
  Amax ~ 1 + water_regime + (1 | plant_id),
  alpha ~ 1 + water_regime,
  Rd ~ 1,
  nl = TRUE
)
pri <- c(
  brms::prior(normal(25, 10), nlpar = "Amax"),
  brms::prior(lognormal(log(0.06), 0.5), nlpar = "alpha", lb = 0),
  brms::prior(normal(1.5, 1), nlpar = "Rd", lb = 0)
)
b <- nl_bayes(bf, p, prior = pri, family = brms::student(), chains = 4, cores = 4)
nl_ppcheck(b, type = "dens_overlay")
```

## Saturation and inhibition models elsewhere in physiology

`nl_models()` also exposes Hill, Emax, Michaelis-Menten and substrate-inhibition equations. These can represent concentration-response, enzyme kinetics or substrate inhibition when their mechanistic assumptions match the experiment.

```{r}
subset(nl_models(), category %in% c("physiology", "dose_response", "saturation"))
```

## Publication output

```{r, eval=FALSE}
pub <- nl_plot(f_nonrect, interval = "confidence")
nl_save_plot(pub, "photosynthetic_light_response.svg", 180, 120, "mm")
nl_table(f_nonrect, "parameters", "gt")
```

```{r eval=FALSE}
nl_bayes_diagnose(b)
# Compare Bayesian light-response specifications after fitting both:
# nl_bayes_compare(list(Rectangular=b_rect, Nonrectangular=b), method="stacking")
```

```{r, eval=FALSE}
# Grid-start search when the light-response model is sensitive to initialization
nl_multistart(A_umol_CO2_m2_s~Amax*alpha*PAR_umol_m2_s/(Amax+alpha*PAR_umol_m2_s)-Rd, ww, c(Amax=5,alpha=.005,Rd=0), c(Amax=60,alpha=.3,Rd=6), iter=c(4,4,4))
```

```{r, eval=FALSE}
# Global-search fallback if the light-response fit remains initialization-sensitive
nl_global_start(A_umol_CO2_m2_s~Amax*alpha*PAR_umol_m2_s/(Amax+alpha*PAR_umol_m2_s)-Rd, ww, c(Amax=1,alpha=.001,Rd=0), c(Amax=80,alpha=.5,Rd=10), pop_size=60, maxiter=300, run=50)
```
