---
title: "Bayesian nonlinear models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian nonlinear models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
library(nlrFlow)
```

Bayesian nonlinear models require explicit, scientifically defensible priors, especially when nonlinear parameters are weakly identified.

```{r eval=FALSE}
# bf <- brms::bf(fruit_length ~ Asym*exp(-exp(-k*(day_after_flowering-xmid))), Asym~1,k~1,xmid~1,nl=TRUE)
# fit <- nl_bayes(bf,nl_data("okra_growth_raw"),prior=...)
# nl_ppcheck(fit)
nl_capabilities()
nl_doctor()
nl_teach("gompertz")
```

```{r, eval=FALSE}
# Sampling diagnostics
nl_bayes_diagnose(fit_bayes)
# PSIS-LOO comparison and stacking weights after fitting alternative models
nl_bayes_compare(list(M1=fit_bayes, M2=fit_bayes_alt), method="stacking")
```


## Posterior intervals and predictions

`nlrFlow` does not apply Wald intervals to Bayesian fits. Parameter summaries use posterior draws, expected-response intervals use `posterior_epred()`, and new-observation intervals use `posterior_predict()`.

```{r eval=FALSE}
nl_confint(fit_bayes, method = "posterior", level = 0.95)

new_okra <- data.frame(day_after_flowering = seq(1, 25, length.out = 100))
nl_predict(fit_bayes, newdata = new_okra,
           interval = "confidence", nsim = 1000)
nl_predict(fit_bayes, newdata = new_okra,
           interval = "prediction", nsim = 1000)
```

For hierarchical models, arguments such as `re_formula` can be passed through `...` to control whether predictions are conditional on, or marginal over, group-level effects.

```{r eval=FALSE}
# Population-level agronomic curve
nl_predict(agri_bayes, newdata = agri_grid,
           interval = "confidence", re_formula = NA)

# Plant-level physiology prediction, retaining fitted group effects
nl_predict(phys_bayes, newdata = phys_grid,
           interval = "prediction", re_formula = NULL)
```

## Publication output

```{r eval=FALSE}
nl_table(fit_bayes, type = "parameters")
nl_plot(fit_bayes,
        x = "day_after_flowering", y = "fruit_length",
        interval = "confidence")
nl_report(fit_bayes, "okra_bayesian_report.md")
```

The plotting function requires explicit `x` and `y` names for `brms` objects because a nonlinear `brmsformula` may contain multiple parameter submodels and cannot be safely reduced to a single base-R formula automatically.

## Posterior biological quantities

Growth rates, inflection timing, midpoint timing and AUC can be propagated through posterior expected-response draws rather than through a frequentist delta approximation.

```{r eval=FALSE}
nl_derive(fit_bayes,
          predictor = "day_after_flowering",
          quantities = c("AGR", "RGR", "inflection", "AGRmax", "t50", "AUC"),
          nsim = 1000,
          re_formula = NA)
```

For treatment contrasts, `nl_contrast()` also accepts a named contrast matrix whose columns match `brms` posterior coefficient names and reports posterior intervals and the posterior probability that each contrast exceeds zero.
