---
title: "Diagnostics and Reproducibility for Scientific Machine Learning"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Diagnostics and Reproducibility for Scientific Machine Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Training loss is not a scientific diagnostic

`nl_sciml_diagnose()` combines trajectory error, available physics-loss information, neural-correction magnitude, training history, and optional repeated-seed fits. SciML's sensitivity infrastructure enables differentiation through differential-equation solvers, but numerical differentiability does not remove the need to test identifiability, stability, and extrapolation [@Rackauckas2020UDE].

# 1. Neural ODE diagnostic cycle

```{r node, eval=FALSE}
fruit <- nl_data("sciml_fruit_growth")
f1 <- subset(fruit,fruit_id==unique(fruit_id)[1])
node <- nl_neural_ode(f1,"day_after_set","fruit_mass_g",c("temperature_C","soil_water_rel"),seed=201)
diag_node <- nl_sciml_diagnose(node)
nl_plot(diag_node)
nl_table(diag_node)
```

# 2. UDE diagnostic cycle

```{r ude, eval=FALSE}
crop <- nl_data("sciml_crop_growth")
c1 <- subset(crop,plot_id==unique(plot_id)[1])
ude <- nl_ude(c1,"day","biomass_g_m2",c("temperature_C","PAR_MJ_m2_d","soil_water_rel","nitrogen_rel"),template="crop_growth_rue",known_parameters=c(RUE=2.8,Topt=26,Twidth=10,respiration=.01),seed=202)
diag_ude <- nl_sciml_diagnose(ude,thresholds=c(rmse_ratio=.20,physics_loss=.05,neural_fraction=.50))
nl_plot(diag_ude)
```

# 3. PINN diagnostic cycle

```{r pinn, eval=FALSE}
sw <- nl_data("sciml_soil_water")
pinn <- nl_pinn(sw,"richards_1d","day","theta_cm3_cm3",depth="depth_cm",covariates=c("rain_mm_d","ET0_mm_d"),parameters=c(theta_r=.06,theta_s=.46,alpha=.035,n=1.55,Ks=8),seed=203)
diag_pinn <- nl_sciml_diagnose(pinn)
nl_table(diag_pinn)
```

# Seed stability

A neural model that changes scientific conclusions across reasonable seeds is not stable enough for mechanistic interpretation. Refit at multiple frozen seeds and pass the fits to `seed_fits`.

```{r seeds, eval=FALSE}
fits <- lapply(c(301,302,303,304,305),function(s) nl_neural_ode(f1,"day_after_set","fruit_mass_g",c("temperature_C","soil_water_rel"),seed=s))
nl_sciml_diagnose(fits[[1]],seed_fits=fits[-1])
```

# Minimum reporting checklist

Report data split and experimental unit, architecture, activation, solver, optimizer sequence, sensitivity method, stopping rule, seeds, trajectory error, physics loss where relevant, parameter bounds, conservation or shape checks, neural-correction magnitude for UDEs, seed sensitivity, interpolation and extrapolation assessment, and a simpler scientific comparator.
